VLI - a library for large integer and polynomial arithemtic

### 2013-01-22 doc/paper_vli/vli2.tex
minor update

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
We present a high-performance C++ library for 128 to 512 bit integer arithmetic and symbolic polynomial computations.
While both parts of the library can be used independently optimized kernels for symbolic polynomials with large integer coefficients are provided.
%Even though the application developed as a special purpose library for high-temperature series expansions, it is also of great interest for other applications due to its modular design.
%The kernels for the large integer operations are implemented in \verb|x86-64| and \verb|power64| assembly.
%The polynomials may have up to 4 symbolic variables and have a compile time fixed truncation order.
%Since our main target applications is high-temperature series expansions (HTSE) we implemented a special routine to perform inner products of vectors of polynomials with large integer coefficients.
Our main target application is high-temperature series expansions (HTSE) which require inner products of large vectors of polynomials with large integer coefficients.
For this purpose we implemented a tunable hybrid CPU/GPU inner product function using OpenMP and NVIDIA CUDA.
This way we make optimal use of todays and upcoming hybrid supercomputers and attain 67\% of the peak performance of the current NVIDIA Kepler GPU.
Compared to a pure CPU solution using the GNU Multiple Precision Arithmetic Library (GMP) we gain a speedup of 10x for a pure CPU inner product and 60x using a GPU accelerator.
\end{abstract}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\section{Large integers}
\begin{itemize}
\item {\bf Description}
\item signed integer arithmetic
\item fixed size (128-512bit)
\item all standard integer operations (arithmetic, comparison and bit operations)
\item in addition to that: extended multiplication which doubles the number of bits, extended add which increases the size by one 64bit.
\item {\bf Implementation details}
\item data is stored as 64 bit integer arrays.
\item stack based $\rightarrow$ cheap memory allocation
\item optimized ASM kernels to use hardware features not accessible from C++ (\verb|x86_64| carry addition \verb|adcq|, 64bit multiplication \verb|mulq| which calculates low and hi 64bit part of the result.)
\item ASM code generated using \verb|BOOST_PP| to compactify the code and minimize bugs.
\end{itemize}

\section{Polynomials}
\begin{itemize}
\item {\bf Description}
\item dense symbolic polynomial in 1-4 variables.
\begin{equation}
    p(x,y,z,a) = \sum_{i,j,k,l} c_{ijkl} x^i y^j z^k a^l
\end{equation}
\item either ``dense'' (${i,j,k,l} \le N$) or ``triangular'' ($i+j+k+l \le N$) structure.
\item any arithmetic type as coefficient $c_{ijkl}$
\item addition, subtraction, multiplication with other polynomials and monomials
\item {\bf Implementation details}
\item structure and truncation order $N$ is compile time fixed 
\item The symbolic variable names are given as template parameter
\item Automatic matching of variable names of different polynomial types using template meta programming.
\end{itemize}

\section{Optimized inner product}
\paragraph{CPU}
\begin{itemize}
\item The hot spot of HTSE are inner products of large vectors with symbolic polynomials as components. The coefficients of the polynomials are large integers.
\item We continue to use the single large integer routines
\item No SIMD, since the number of operations for a multiplication increases by a factor TODO due to the manual handling of the carry and separate lo/hi word multiplication.
\item OpenMP parallelization
\end{itemize}
\paragraph{GPU}
\begin{itemize}
\item Many small tasks of the same form. $\rightarrow$ well suited for GPUs.
\item {\bf Implementation details}
\item Each resulting coefficient of the polynomial $c_{I,J,K,L}$ may has several e
\end{itemize}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
We present a high-performance C++ library for 128 to 512 bit integer arithmetic and symbolic polynomial computations.
While both parts of the library can be used independently optimized kernels for symbolic polynomials with large integer coefficients are provided.
Our main target application is high-temperature series expansions (HTSE) which require inner products of large vectors of polynomials with large integer coefficients.
For this purpose we implemented a tunable hybrid CPU/GPU inner product function using OpenMP and NVIDIA CUDA.
This way we make optimal use of todays and upcoming hybrid supercomputers and attain 67\% of the peak performance of the current NVIDIA Kepler GPU.
Compared to a pure CPU solution using the GNU Multiple Precision Arithmetic Library (GMP) we gain a speedup of 10x for a pure CPU inner product and 60x using a GPU accelerator.
\end{abstract}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\section{Large integers}
\begin{itemize}
\item {\bf Description}
\item signed integer arithmetic
\item fixed size (128-512bit)
\item all standard integer operations (arithmetic, comparison and bit operations)
\item in addition to that: fused multiply-add, extended multiplication which doubles the number of bits, extended add which increases the size by one 64bit.
\item {\bf Implementation details}
\item data is stored as 64 bit integer arrays.
\item stack based $\rightarrow$ cheap memory allocation
\item optimized ASM kernels to use hardware features not accessible from C++ (\verb|x86_64| carry addition \verb|adcq|, 64bit multiplication \verb|mulq| which calculates low and hi 64bit part of the result.)
\item ASM code generated using \verb|BOOST_PP| to compactify the code and minimize bugs.
\end{itemize}

\section{Polynomials}
\begin{itemize}
\item {\bf Description}
\item symbolic polynomial in 1-4 variables.
\begin{equation}
    p(x,y,z,a) = \sum_{i,j,k,l} c_{ijkl} x^i y^j z^k a^l
\end{equation}
\item either ``dense'' (${i,j,k,l} \le N$) or ``triangular'' ($i+j+k+l \le N$) structure.
\item any arithmetic type as coefficient $c_{ijkl}$
\item addition, subtraction, multiplication with other polynomials and monomials
\item {\bf Implementation details}
\item structure and truncation order $N$ is compile time fixed 
\item The symbolic variable names are given as template parameter
\item Automatic matching of variable names of different polynomial types using template meta programming.
\end{itemize}

\section{Optimized inner product}
\paragraph{CPU}
\begin{itemize}
\item The hot spot of HTSE are inner products of large vectors with symbolic polynomials as components. The coefficients of the polynomials are large integers.
\item We continue to use the single large integer routines
\item No SIMD, since the number of operations for a multiplication increases by a factor TODO due to the manual handling of the carry and separate lo/hi word multiplication.
\item OpenMP parallelization
\end{itemize}
\paragraph{GPU}
\begin{itemize}
\item Many small tasks of the same form. $\rightarrow$ well suited for GPUs.
\item {\bf Implementation details}
\item Each resulting coefficient of the polynomial $c_{IJKL}$ may has several e
\end{itemize}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
